Future cosmic microwave background delensing with galaxy surveys.

### 2017-10-26 Notes/galaxy_delensing.tex
more text review also after Giulio input

Because the CMB B-modes provide the cleanest known observational window into the primordial gravitational wave background, improving their measurement is a major objective of current and future CMB experiments.
Even if inflationary B-modes have not been detected yet, the target value $r \gtrsim 10^{-3}$, should be reachable shortly given the level of noise expected in future CMB experiments \citep{kamionkowski:2016,abazajian:2016}.
However, just reducing the level of noise will not be enough to attain this goal.
Indeed, the observed B modes are not solely sourced by early universe physics; they are also produced by secondary effects taking place in the late-time low-redshift universe. 
In this work, we will focus on the effect of gravitational interactions with large scale structures (LSS).
Lensing shears the CMB polarization pattern, producing ``lensing B modes'' from CMB E modes \citep{zaldarriaga98}.
This expected component has now been measured both in cross-correlation with LSS \citep{hanson13,bicep2-collaboration:2016,polarbear2014c,van-engelen:2015,planck2015XV} and from CMB data alone \citep{polarbear2014b,bicep2a,keisler15}.

@@

The optimal way to reduce the lensing contributions to the B-modes is to reconstruct a template of the actual lensing B modes (or the actual large scale structure lensing potential) on the observed part of the sky and then use it to clean the data in a process called ``delensing''. 
We can delens the observed B-modes by combining CMB polarization data (what is \textit{lensed}) with tracers of the large scale structure (what is \textit{lensing}) to reconstruct a template of the expected lensing B-modes.
Delensing has been studied for many years \cite{knox2002, kesden2002, seljak2003, simard:2015,sherwin15,smith:2012}.
Furthermore, it has recently been performed on CMB temperature data using the cosmic infrared background (CIB) as LSS tracer \cite{larsen:2016a} and on CMB temperature and polarization data using CMB data to internally reconstruct the LSS lensing potential \cite{carron:2017}. Finally, the highest B-mode delensing efficiency has been achieved with SPT and Herschel data in which $~28\%$ of the lensing power was removed \cite{manzotti:2017}.

@@

\section{Lensing potential tracers}
\label{sec:model}
In this section, we introduce the different large scale structure tracers considered in this work to reconstruct the lensing potential. Gravitational lensing distorts primordial E-modes generating a non-primordial B-mode component.
Also, we define the power spectra that we will use later in \refsec{th}.

Large scale structure surveys usually probe the 3D matter overdensities as a 2D field projected along the line of sight:
\begin{equation}
\delta^{i}(\nver) = \int_0^{\infty} dz\, W^{i}(z)\delta(\chi(z)\nver,z).
\label{eqn:wkernel}
\end{equation}
where $\delta(\chi(z)\nver,z)$ corresponds to the dark matter overdensity field at a comoving distance $\chi(z)$ and at a redshift $z$ in the angular direction $\nver$.
Using the Limber approximation \cite{limber53} we can compute the power spectra of two large-scale structure fields $i,j$ as:
\begin{equation}
C_{\ell}^{ij}= \int_0^{\infty} \frac{dz}{c} \frac{H(z)}{\chi(z)^2}\, W^{i}(z)W^{j}(z)P(k,z).
\label{eqn:wkappa}{}
\end{equation}

In this equation, $H(z)$ is the Hubble factor at redshift $z$, $c$ is the speed of light, and $P(k,z)$ is the matter power spectrum evaluated at wavenumber $k = \ell/\chi(z)$ and redshift $z$. 
%Furthermore $W^{i}(z)$ is the kernel function of the field $i$ introduced in \refeq{wkernel}.
We will now describe the kernels $W^{i}(z)$ for each of the tracers used in this work.

@@

The galaxy clustering kernel is:
\begin{equation}
\label{eqn:wg}
\begin{split}
W^{g}(z) &= \frac{b(z)\frac{dN}{dz}}{\Bigl(\int dz'\,\frac{dN}{dz'}\Bigr)}.
\end{split}
\end{equation}
Here $\frac{dN}{dz}$ is the number of galaxies observed by the survey as a function of redshift while $b(z)$ is the galaxy bias that connects the amplitude of galaxy overdensities to the underlying dark matter density. 
We use a linear bias independent of the angular scale, which is a reasonable assumption for the relatively large scales relevant for delensing ($\ell<1000$).
When computing the auto-spectrum of the galaxy density, a shot noise term needs to be taken into account. 
To do so, we add a constant term to the power spectrum equal to the inverse of the number of galaxies per steradians.
 Different galaxy surveys in this work are then fully characterized by their b(z), $\frac{dN}{dz}$ and the observed galaxy density.
We test the delensing efficiency taking into account both current surveys like WISE or DES as well as future galaxy surveys like DESI and LSST as well as 21 cm measurement like SKA.

@@

There are several available CIB observations that have already been used to delens the CMB. Given its full-sky coverage, a promising one is the CIB map derived in \cite{2016A&A...596A.109P} using multi-frequency Planck data and the Generalized Needlet Internal Linear Combination (GNILC) component separation algorithm.

@@

Indeed the Q and U mode decompositions of the CMB photons polarization are remapped by lensing as:
\be
Q(\hat{\mathbf{n}}) = Q_{\mathrm{unlensed}}(\hat{\mathbf{n}}+\nabla\phi);~~
U(\hat{\mathbf{n}}) = U_{\mathrm{unlensed}}(\hat{\mathbf{n}}+\nabla\phi)
\ee
where the deflection angle is the gradient of the lensing potential integrated along the line of sight $\nabla\phi$.
The CMB polarization is usually decomposed into odd-parity Fourier modes E and B. 
As shown in \cite{seljak:1997}, because of the symmetry of the problem, tensor perturbations are the principal source of the B-modes configuration. For this reason, B-modes are a promising signature of early universe tensor perturbations.

However, this promising sign of primordial gravitational waves is partially obscured by gravitational interactions with large scale structures that generate CMB B-modes by distorting primordial E-modes.
At first order, given the convergence field $\kappa= -\frac{1}{2}\nabla^2\phi$ introduced in \refsec{kappaCMB} the B-modes resulting from the lensing of the primordial E-modes are:
\be
B^{\mathrm{lens}}(\bl) =  \int \frac{d^2 \bl'}{(2 \pi)^2} W(\bl,\bl') E(\ublu) \kappa(\bl - \bl')
\label{eqn:blens}
\ee
where different modes contributes with a different weight:
\be
W(\bl,\bl') = \frac{2 \bl' \cdot (\bl-\bl')}{|\bl-\bl'|^2} \sin(2\varphi_{\bl,\bl'}).
\ee
Here $\varphi_{\bl,\bl'}$ is the angle between the two different modes $\bl$ and $\bl'$.
From this we get the power spectrum of the lensing component of the B-modes:
\be
C_{\ell}^{BB,\mathrm{lens}}  = \int \frac{d^2 \bl'}{(2 \pi)^2} W^2 (\bl,\bl')C^{EE}_{l'} C^{\kappa \kappa}_{|\bl-\bl'|}.
\ee

The B-mode power spectrum measured on the sky is composed of a possible primordial component $C_{\ell}^{BB,r}$ together with the lensing $C_{\ell}^{BB,\mathrm{lens}}$ contribution and the instrumental noise $N_{\ell}^{BB}$ (defined in \refeq{noise}):
\be
C_{\ell}^{BB,\mathrm{measured}} = C_{\ell}^{BB,r} + C_{\ell}^{BB,\mathrm{lens}} + N_{\ell}^{BB}.
\ee
The lensing component is a significant source of B-modes that, at large scales, corresponds to a white noise source of roughly $5 \mu K$-arcmin independent of the angular scale. This means that it is not only larger than the allowed inflationary component at scales smaller than several degrees, but it is also comparable to current levels of instrumental noise. 
For this reason, it is critical to characterize and eventually remove it from the data.
To do so, we build a template \refeq{blens} of the lensing B-modes in the observed patch given a measurement of the E-mode field and the lensing potential $\phi$.
While E is measured directly, we can estimate $\phi$ using "tracers" of the matter distribution that sources the potential.

@@

The optimal weights $f(\bl,\bl')$, chosen such that the residual lensing B mode power is minimized, are:
\be
\label{eqn:fweight}
f(\bl,\bl') = \left(\frac{C^{EE}_{{l'}}}{C^{EE}_{{l'}}+N^{EE}_{{l'}}}\right)  \frac{C^{\kappa I}_{|\bl-\bl'|}}{C^{II}_{|\bl-\bl'|} }.
\ee
Here $C^{\kappa I}$ and $C^{II}$ are the cross-correlation spectrum of the tracer $I$ with the lensing convergence $\kappa$ and its autospectrum; they are described for each LSS field in \refsec{model}. The power spectrum of the E-modes noise $N^{EE}$ is the same as the B-modes one in \refeq{noise}.

@@

% ===================
\begin{figure}[htbp]
\begin{center}
\includegraphics[scale=1.]{../images/compare_kernel.pdf}
\caption{Kernel Comparison: Comparison of the different kernels as a function of redshift for some of the tracers used in this analysis. The bigger the overlap with the CMB lensing kernel the better the reconstruction of the lensing potential is thus leading to a higher delensing efficiency. }
\label{fig:kernel}
\end{center}
\end{figure}
% ===================

@@

We start by assuming that we have n different tracers of the gravitational potentials $I_{i}$ with $i\in \{1,..,n\}$. We can optimally combine them to estimate $\phi$ or, in other words, to maximize the correlation factor $\rho$ with:
\beqn
I         &=&    \sum_{i}c^{i}I^{i} \nonumber \\
c_{i}     &=& (C_{II}^{-1})_{ij}C^{{\kappa I}^{j}}
\label{eqn:combined}
\eeqn
where $C_{II}$ is the covariance matrix of the LSS tracers.
The residual B-mode power can be derived from \refeq{Bres} using an ``effective'' correlation $\rho^{2}$ of these combined tracers with gravitational lensing:
\be
\rho^{2}_{\ell} = \sum_{i,j}\frac{C^{\kappa I^{i}}_{\ell}~(C^{II}_{\ell})^{-1}_{ij}~C^{\kappa I^{j}}_{\ell}}{C^{\kappa\kappa}_{\ell}}.
\label{eqn:rho-combined}
\ee

Note that the gain in adding a new tracer is not only proportional to its correlation with the CMB lensing, but it also depends on how much it is correlated with the already used set of tracers.
\reffig{kernel} show the different kernels as a function of redshift computed using the models and parameters described in \refsec{model}. The cross-correlation of a tracer with the CMB lensing is directly proportional to the overlap of their kernels. 
The cosmic infrared background and 21 cm surveys probe the high redshift structures and, independent of the model assumed, they show a relatively good overlap with the CMB lensing kernel. 
On the other end, galaxy clustering surveys can only reconstruct the low-z portion of the lensing kernel as can be seen from the LSST, DES and DESI curves.
However, given the low noise of these measurements and their small overlap with other probes, they can still play a major role in delensing even if their overlap with the CMB lensing potential is not optimal.

\subsection{Improving efficiency with tomographic binning}\label{sec:tomo-bin}
The delensing efficiency of galaxy surveys can be improved by taking into account redshift information.
When we weight a tracer with $\frac{C^{\kappa I}}{C^{II}}$ in \refeq{fweight} in order to maximize its ability to reconstruct the lensing potential we are only using redshift averaged information about the survey. However, as can be seen in \reffig{kernel}, the kernel overlap of a tracer with the CMB lensing varies as a function of redshift. For this reason, the optimal approach is to weight galaxies at different redshift with different weights according to both their cross-correlation with $\kappa$ and their auto-spectrum.
We can see this with a simple example. Let's split a single survey I into two non-overlapping redshift bins $I_{1}$ and ${I_{2}}$ with $I=I_{1}+I_{2}$.
For the full survey the effective cross-correlation is equal to 
\be
\rho_{\rm{full}}^{2}= \frac{(\cl^{\kappa I})^2}{\cl^{\kappa \kappa} \cl^{I I}}=\frac{(\cl^{\kappa I_{1}}+\cl^{\kappa I_{2}})^2}{\cl^{\kappa \kappa} (\cl^{I_{1} I_{1}}+\cl^{I_{2} I_{2}})}
\ee
while for the split survey it will be 
\be
\rho_{\rm{split}}^{2}= \frac{(\cl^{\kappa I_{1}})^2}{\cl^{\kappa \kappa} \cl^{I_{1} I_{1}}}+\frac{(\cl^{\kappa I_{2}})^2}{\cl^{\kappa \kappa} \cl^{I_{2} I_{2}}}.
\ee
Now it can be show that $\rho_{\rm{split}}\ge \rho_{\rm{full}}$ since:
\be
\rho_{\rm{split}}^{2} - \rho_{\rm{full}}^{2}\propto  (\cl^{I_{1} I_{1}}\cl^{\kappa I_{2}}-\cl^{I_{2} I_{2}}\cl^{\kappa I_{1}})^2
\ee
Then $\rho^2$ is always larger in the tomographic case, and the two are equal only when $\frac{C^{\kappa I_{i}}}{C^{I_{i}I_{i}}}$ is the same for all the redshift bins in which case a single optimal weight is sufficient for the entire survey. 
Binning will improve the efficiency of galaxy surveys, but it is not very effective for tracers with poor redshift information like the CIB or radio continuum surveys.

In this work, we bin both photometric and spectroscopic galaxy surveys by splitting the window function \refeq{wg} into different slices such as all the bins contain the same number of galaxies. 
For photometric surveys like DES and LSST, we assume a photometric redshift estimation gaussianly distributed around the true value with an rms fluctuation $\sigma(z)$.

@@

To quantify the improvement after delensing we introduce the error on a parameter $p$ constrained with \textit{delensed} spectra as $\sigma^{\rm{del}}(p)$.
This is computed from \refeq{fisher} substituting $C_{{l}}^{BB,\mathrm{lens}}$ with the residual power after delensing $C_{{l}}^{BB,\mathrm{res}}$ in the covariance matrix defined in \refeq{cova}.
The improvement is then the ratio of the constraints before and after delensing: $\alpha_{r} = \sigma^{\rm{del}}(r)/\sigma(r)$ and $\alpha_{n_{t}} = \sigma^{\rm{del}}(n_{t})/\sigma(n_{t})$. 
The ratio $\alpha_{r,n_{t}}$ is insensitive to the fraction of the sky observed by the experiment: while the absolute parameter constraints depend on $\fsky$ the relative improvement does not.  For this reason, we do not quote $\fsky$ values for the CMB experiments investigated here.

@@

%===========================
%===========================
\begin{table}
\caption{$\alpha(r)$: Improvements on $\sigma(r)$ due to delensing for completed and ongoing Stage-2 surveys and Planck lensing reconstruction.
LSS-S2 corresponds to the combination of all the available LSS tracers.
The values in parenthesis in the first column correspond to the fraction of lensing B-mode power removed using each LSS tracer. 
It is defined as $\left<C_{\ell}^{BB,\mathrm{res}}/C_{l}^{BB,\mathrm{lens}}\right>_{4<\ell<100}$.
%\newline
The values in the other columns correspond to ratio of the error before and after delensing for 3 cases: no instrumental noise in the B-mode measurement (but with instrumental noise in the E-mode used to delens) and realistic noise for two different values of $r$.}
  \vspace{0.2cm}
  \begin{tabular}{|c | c | c | c|}
\hline
Surveys & $\alpha_{r=0}, N^{B}_{\ell}=0$ & $\alpha_{r=0}$ & $\alpha_{r=0.12}$ \\ \hline \hline
WISE (6\%)& 1.09& 1.06 & 1.02 \\ \hline
DES (11\%) &1.16 & 1.10  & 1.04 \\ \hline
CIB (25\%) &1.37& 1.21  & 1.09\\ \hline
LSS-S2 (34\%) &1.53& 1.29  & 1.11\\ \hline
CMB Planck (8\%) &1.08& 1.05  & 1.02 \\ \hline
LSS-S2 +CMB (37\%)& 1.58& 1.33& 1.13  \\ \hline
\end{tabular}
\label{tab:current}
\end{table}
%===========================
%===========================

@@

 \begin{table}
\caption{$\alpha(r)$: Stage-2 improvements on $\sigma(r)$ due to delensing. We use the same LSS tracers as \reftab{current} but with the internal lensing reconstruction performed by SPTPol.}
  \vspace{0.2cm}
  \begin{tabular}{|c | c | c | c|}
\hline
Surveys &$\alpha_{r=0}, N^{B}_{\ell}=0$& $\alpha_{r=0}$ & $\alpha_{r=0.12}$ \\ \hline \hline
SPTPol (35\%) &1.46& 1.27  & 1.08 \\ \hline
LSS-S2 +SPTPol (58\%) & 2.18 & 1.57 & 1.16  \\ \hline

@@

\begin{table}
\caption{$\alpha$: Improvements on $\sigma(r)$ due to delensing for S3 experiments. Here we define LSS-S3 as the combination of S2 LSS tracers (LSS-S2) and DESI. We also use a CMB S3 experiment for the lensing internal reconstruction. See \refsec{S3} for details.}
%The absolute error value of $\sigma(r)$ after delensing for $f_{sky} = $ is $\sigma(r) =$ 
\hspace{2.2cm}
\begin{tabular}{|c | c | c | c|}
\hline
Surveys & $\alpha_{r=0}, N^{B}_{\ell}=0$ & $\alpha_{r=0}$ & $\alpha_{r=0.12}$ \\ \hline \hline
DESI (10\%) &1.12 & 1.08  & 1.03 \\ \hline
LSS-S3 (40\%) &1.73& 1.42  & 1.2\\ \hline
CMB 3G (56\%) &2& 1.57  & 1.21 \\ \hline
LSS-S3+CMB (69\%)& 2.8 & 1.8& 1.31  \\ \hline
\end{tabular}
\label{tab:S3}
\end{table}

>>>>

Because the CMB B-modes provide the cleanest known observational window into the primordial gravitational wave background, improving their measurement is a major objective of current and future CMB experiments.
Even if inflationary B-modes have not been detected yet, the target value $r \gtrsim 10^{-3}$, should be reachable shortly given the level of noise expected in future CMB experiments \citep{kamionkowski:2016,abazajian:2016}.
However, just reducing the level of noise will not be enough to reach this target.
Indeed, the observed B modes are not solely sourced by early universe physics; they are also produced by secondary effects taking place in the late-time low-redshift universe. 
In this work, we will focus on the effect of gravitational interactions with large scale structures (LSS).
Lensing shears the CMB polarization pattern, producing ``lensing B modes'' from CMB E modes \citep{zaldarriaga98}.
This expected component has now been measured both in cross-correlation with LSS \citep{hanson13,bicep2-collaboration:2016,polarbear2014c,van-engelen:2015,planck2015XV} and from CMB data alone \citep{polarbear2014b,bicep2a,keisler15}.

@@

The optimal way to reduce the lensing contributions to the B-modes is to reconstruct a template of the actual lensing B modes (or the actual large scale structure lensing potential) on the observed part of the sky and then use it to marginalize the lensing contribution in the data in a process called ``delensing''. 
We can delens the observed B-modes by combining CMB polarization data (what is \textit{lensed}) with tracers of the large scale structure (what is \textit{lensing}) to reconstruct a template of the expected lensing B-modes.
Delensing has been studied for many years \cite{knox2002, kesden2002, seljak2003, simard:2015,sherwin15,smith:2012}.
Furthermore, it has recently been performed on CMB temperature data using the cosmic infrared background (CIB) as LSS tracer \cite{larsen:2016a} and on CMB temperature and polarization data using CMB data to internally reconstruct the LSS lensing potential \cite{carron:2017}. Finally, the highest B-mode delensing efficiency has been achieved with SPT and Herschel data in which $~28\%$ of the lensing power was removed \cite{manzotti:2017}.

@@

\section{Lensing potential tracers}
\label{sec:model}
In this section, we introduce the different large scale structure tracers considered in this work to reconstruct the lensing potential. We will see how large scale structures distorts primordial E-modes generating a non-primordial B-mode component in the next section.
Also, we define the power spectra that we will use later in \refsec{th}.

Large scale structure surveys usually probe the 3D matter overdensities as a 2D field projected along the line of sight:
\begin{equation}
\delta^{i}(\nver) = \int_0^{\infty} dz\, W^{i}(z)\delta(\chi(z)\nver,z).
\label{eqn:wkernel}
\end{equation}
where $\delta(\chi(z)\nver,z)$ corresponds to the dark matter overdensity field at a comoving distance $\chi(z)$ and at a redshift $z$ in the angular direction $\nver$.
Using the Limber approximation \cite{limber53} we can compute the power spectra of two large-scale structure fields $i,j$ as:
\begin{equation}
C_{\ell}^{ij}= \int_0^{\infty} \frac{dz}{c} \frac{H(z)}{\chi(z)^2}\, W^{i}(z)W^{j}(z)P(k = \ell/\chi(z),z).
\label{eqn:wkappa}{}
\end{equation}

In this equation, $H(z)$ is the Hubble factor at redshift $z$, $c$ is the speed of light, and $P(k,z)$ is the matter power spectrum evaluated at redshift $z$. 
%Furthermore $W^{i}(z)$ is the kernel function of the field $i$ introduced in \refeq{wkernel}.
We will now describe the kernels $W^{i}(z)$ for each of the tracers used in this work.

@@

The normalized galaxy clustering kernel is:
\begin{equation}
\label{eqn:wg}
\begin{split}
W^{g}(z) &= \frac{b(z)\frac{dN}{dz}}{\Bigl(\int dz'\,\frac{dN}{dz'}\Bigr)}.
\end{split}
\end{equation}
Here $\frac{dN}{dz}$ is the number of galaxies observed by the survey as a function of redshift while $b(z)$ is the galaxy bias that connects the amplitude of galaxy overdensities to the underlying dark matter density. 
We use a linear bias independent of the angular scale, which is a reasonable assumption for the relatively large scales relevant for delensing ($\ell<1000$).
When computing the auto-spectrum of the galaxy density, a shot noise term needs to be taken into account. 
To do so, we add a constant term to the power spectrum equal to the inverse of the number of galaxies per steradians.
 Different galaxy surveys in this work are then fully characterized by their $b(z)$, $\frac{dN}{dz}$ and the observed galaxy density.
We test the delensing efficiency taking into account both current surveys like WISE or DES as well as future galaxy surveys like DESI and LSST as well as 21 cm measurement like SKA.

@@

There are several available CIB observations that have already been used to delens the CMB. Given its full-sky coverage and the small contribution from foregrounds contamination, a promising CIB map for future experiments is the one derived in \cite{2016A&A...596A.109P} using multi-frequency Planck data and the Generalized Needlet Internal Linear Combination (GNILC) component separation algorithm.

@@

Indeed the Q and U mode decompositions of the CMB photons polarization are remapped by lensing as:
\be
Q(\hat{\mathbf{n}}) = Q_{\mathrm{unlensed}}(\hat{\mathbf{n}}+\nabla\phi);~~
U(\hat{\mathbf{n}}) = U_{\mathrm{unlensed}}(\hat{\mathbf{n}}+\nabla\phi)
\ee
where the deflection angle is the gradient of the lensing potential integrated along the line of sight $\nabla\phi$.
The CMB polarization is usually decomposed into odd-parity Fourier modes E and B. 
As shown in \cite{seljak:1997}, because of the symmetry of the problem, tensor perturbations are the principal source of the B-modes configuration. For this reason, primordial B-modes are a promising signature of early universe tensor perturbations.

However, primordial B-modes are obscured by gravitational interactions between the large scale structures and the CMB that generate CMB B-modes by distorting primordial E-modes.
At first order, given the convergence field $\kappa= -\frac{1}{2}\nabla^2\phi$ introduced in \refsec{kappaCMB} the B-modes resulting from the lensing of the primordial E-modes are:
\be
B^{\mathrm{lens}}(\bl) =  \int \frac{d^2 \bl'}{(2 \pi)^2} W(\bl,\bl') E(\ublu) \kappa(\bl - \bl')
\label{eqn:blens}
\ee
where different modes contributes with a different weight:
\be
W(\bl,\bl') = \frac{2 \bl' \cdot (\bl-\bl')}{|\bl-\bl'|^2} \sin(2\varphi_{\bl,\bl'}).
\ee
Here $\varphi_{\bl,\bl'}$ is the angle between the two different modes $\bl$ and $\bl'$.
From this we get the power spectrum of the lensing component of the B-modes:
\be
C_{\ell}^{BB,\mathrm{lens}}  = \int \frac{d^2 \bl'}{(2 \pi)^2} W^2 (\bl,\bl')C^{EE}_{l'} C^{\kappa \kappa}_{|\bl-\bl'|}.
\ee

The B-mode power spectrum measured on the sky is composed of a possible primordial component $C_{\ell}^{BB,r}$ together with the lensing $C_{\ell}^{BB,\mathrm{lens}}$ contribution and the instrumental noise $N_{\ell}^{BB}$ (defined in \refeq{noise}):
\be
C_{\ell}^{BB,\mathrm{measured}} = C_{\ell}^{BB,r} + C_{\ell}^{BB,\mathrm{lens}} + N_{\ell}^{BB}.
\ee
The lensing component is a significant source of B-modes that, at large scales, corresponds to a white noise source of roughly $5 \mu K$-arcmin independent of the angular scale. This means that it is not only larger than the allowed inflationary component at scales smaller than several degrees ($r_{0.05} < 0.12$ from \cite{bicep2/keck-collaboration:2015}), but it is also comparable to current levels of instrumental noise. 
For this reason, it is critical to characterize and eventually remove it from the data.
To do so, while other approaches are possible (\cite{larsen:2016a,carron:2017,millea:2017}), here we assume a "template approach": we build a template \refeq{blens} of the lensing B-modes in the observed patch given a measurement of the E-mode field and the lensing potential $\phi$.
While E is measured directly, we can estimate $\phi$ using "tracers" of the matter distribution that sources the potential.

@@

The optimal weights $f(\bl,\bl')$, chosen such that the residual lensing B mode power is minimized, are \cite{sherwin:2015}:
\be
\label{eqn:fweight}
f(\bl,\bl') = \left(\frac{C^{EE}_{{l'}}}{C^{EE}_{{l'}}+N^{EE}_{{l'}}}\right)  \frac{C^{\kappa I}_{|\bl-\bl'|}}{C^{II}_{|\bl-\bl'|} }.
\ee
Here $C^{\kappa I}$ and $C^{II}$ are the cross-correlation spectrum of the tracer $I$ with the lensing convergence $\kappa$ and its autospectrum; they are described for each LSS field in \refsec{model}. The power spectrum of the E-modes noise $N^{EE}$ is the same as the B-modes one in \refeq{noise}.

@@

% ===================
\begin{figure}[htbp]
\begin{center}
\includegraphics[scale=1.]{../images/compare_kernel.pdf}
\caption{Kernel Comparison: Comparison of the different kernels as a function of redshift for some of the tracers used in this analysis. The bigger the overlap with the CMB lensing kernel the better the reconstruction of the lensing potential is thus leading to a higher delensing efficiency. Note that for galaxy surveys the kernel corresponds to $b(z)\frac{dN}{dz}$ and not simply to the redshift distribution of galaxies.}
\label{fig:kernel}
\end{center}
\end{figure}
% ===================

@@

We start by assuming that we have n different tracers of the gravitational potentials $I_{i}$ with $i\in \{1,..,n\}$. We can optimally combine them to estimate $\phi$ or, in other words, to maximize the correlation factor $\rho$ with \cite{sherwin:2015}:
\beqn
I         &=&    \sum_{i}c^{i}I^{i} \nonumber \\
c_{i}     &=& (C_{II}^{-1})_{ij}C^{{\kappa I}^{j}}
\label{eqn:combined}
\eeqn
where $C_{II}$ is the covariance matrix of the LSS tracers. This is assumed to be gaussian and it is computed using \refeq{wkappa}.
The residual B-mode power can be derived from \refeq{Bres} using an ``effective'' correlation $\rho^{2}$ of these combined tracers with gravitational lensing:
\be
\rho^{2}_{\ell} = \sum_{i,j}\frac{C^{\kappa I^{i}}_{\ell}~(C^{II}_{\ell})^{-1}_{ij}~C^{\kappa I^{j}}_{\ell}}{C^{\kappa\kappa}_{\ell}}.
\label{eqn:rho-combined}
\ee

Note that the gain in adding a new tracer is not only proportional to its correlation with the CMB lensing, but it also depends on how much it is correlated with the already used set of tracers.
\reffig{kernel} show the different kernels as a function of redshift computed using the models and parameters described in \refsec{model}. The cross-correlation of a tracer with the CMB lensing is directly proportional to the overlap of their kernels. 
It can be seen, for example, that the cosmic infrared background and 21 cm surveys probe the high redshift structures and, independent of the model assumed, they show a relatively good overlap with the CMB lensing kernel. 
On the other end, galaxy clustering surveys can only reconstruct the low-z portion of the lensing kernel as can be seen from the LSST, DES and DESI curves.
However, given the low noise of these measurements and their small overlap with other probes, they can still play a major role in delensing even if their overlap with the CMB lensing potential is not optimal.

\subsection{Improving efficiency with tomographic binning}\label{sec:tomo-bin}
The delensing efficiency of galaxy surveys can be improved by taking into account redshift information.
When we weight a tracer with $\frac{C^{\kappa I}}{C^{II}}$ in \refeq{fweight} in order to maximize its ability to reconstruct the lensing potential we are only using redshift averaged information about the survey. However, as can be seen in \reffig{kernel}, the kernel overlap of a tracer with the CMB lensing varies as a function of redshift and the latter decreases steeply at low redshift. For this reason, the optimal approach is to weight galaxies at different redshift with different weights according to both their cross-correlation with $\kappa$ and their auto-spectrum.
We can see this with a simple example. Let's split a single survey I into two non-overlapping redshift bins $I_{1}$ and ${I_{2}}$ with $I=I_{1}+I_{2}$.
For the full survey the effective cross-correlation is equal to 
\be
\rho_{\rm{full}}^{2}= \frac{(\cl^{\kappa I})^2}{\cl^{\kappa \kappa} \cl^{I I}}=\frac{(\cl^{\kappa I_{1}}+\cl^{\kappa I_{2}})^2}{\cl^{\kappa \kappa} (\cl^{I_{1} I_{1}}+\cl^{I_{2} I_{2}})}
\ee
while for the split survey it will be 
\be
\rho_{\rm{split}}^{2}= \frac{(\cl^{\kappa I_{1}})^2}{\cl^{\kappa \kappa} \cl^{I_{1} I_{1}}}+\frac{(\cl^{\kappa I_{2}})^2}{\cl^{\kappa \kappa} \cl^{I_{2} I_{2}}}.
\ee
Now it can be show that $\rho_{\rm{split}}\ge \rho_{\rm{full}}$ since:
\be
\rho_{\rm{split}}^{2} - \rho_{\rm{full}}^{2}\propto  (\cl^{I_{1} I_{1}}\cl^{\kappa I_{2}}-\cl^{I_{2} I_{2}}\cl^{\kappa I_{1}})^2
\ee
Then $\rho^2$ is always larger in the tomographic case, and the two are equal only when $\frac{C^{\kappa I_{i}}}{C^{I_{i}I_{i}}}$ is the same for all the redshift bins in which case a single optimal weight is sufficient for the entire survey. 
Binning will improve the efficiency of galaxy surveys, but it is not very effective for tracers with poor redshift information like the CIB or radio continuum surveys.

In this work, we bin both photometric and spectroscopic galaxy surveys by splitting the window function \refeq{wg} into different slices such as all the bins contain the same number of galaxies. For photometric surveys like DES and LSST, we assume a photometric redshift estimation Gaussianly distributed around the true value with an rms fluctuation $\sigma(z)$.

@@

To quantify the improvement after delensing we introduce the error on a parameter $p$ constrained with \textit{delensed} spectra as $\sigma^{\rm{del}}(p)$.
Motivated by \cite{namikawa:2015} we assume the likelihood after delensing is well approximated by a multivariate gaussian. Then we compute $\sigma^{\rm{del}}(p)$ from \refeq{fisher} substituting $C_{{l}}^{BB,\mathrm{lens}}$ with the residual power after delensing $C_{{l}}^{BB,\mathrm{res}}$ in the covariance matrix defined in \refeq{cova}.
The improvement is then the ratio of the constraints before and after delensing: $\alpha_{r} = \sigma^{\rm{del}}(r)/\sigma(r)$ and $\alpha_{n_{t}} = \sigma^{\rm{del}}(n_{t})/\sigma(n_{t})$. 
The ratio $\alpha_{r,n_{t}}$ is insensitive to the fraction of the sky observed by the experiment: while the absolute parameter constraints depend on $\fsky$ the relative improvement does not. For this reason, we do not quote $\fsky$ values for the CMB experiments investigated here.

@@

%===========================
%===========================
\begin{table}
\caption{$\alpha(r)$: Improvements on $\sigma(r)$ due to delensing for completed and ongoing Stage-2 surveys and Planck lensing reconstruction.
LSS-S2 corresponds to the combination of all the available LSS tracers.
The values in parenthesis in the first column correspond to the fraction of lensing B-mode power removed using each LSS tracer. 
It is defined as $\left<C_{\ell}^{BB,\mathrm{res}}/C_{l}^{BB,\mathrm{lens}}\right>_{4<\ell<100}$.
%\newline
The values in the other columns correspond to ratio of the error before and after delensing for 3 cases: no instrumental noise in the B-mode measurement (but with instrumental noise in the E-mode used to delens) and realistic noise for two different values of $r$.}
  \vspace{0.2cm}
  \begin{tabular}{|c | c | c | c|}
\hline
Surveys & $\alpha_{r=0}, N^{B}_{\ell}=0$ & $\alpha_{r=0}$ & $\alpha_{r=0.12}$ \\ \hline \hline
WISE (7\%)& 1.09& 1.06 & 1.02 \\ \hline
DES (15\%) &1.19 & 1.12  & 1.04 \\ \hline
CIB (31\%) &1.48& 1.28  & 1.08\\ \hline
LSS-S2 (41\%) &1.76& 1.41  & 1.11\\ \hline
CMB Planck (7\%) &1.08 & 1.05  & 1.02 \\ \hline
LSS-S2 +CMB (44\%)& 1.83& 1.44& 1.13  \\ \hline
\end{tabular}
\label{tab:current}
\end{table}
%===========================
%===========================

@@

 \begin{table}
\caption{$\alpha(r)$: Stage-2 improvements on $\sigma(r)$ due to delensing. We use the same LSS tracers as \reftab{current} but with the internal lensing reconstruction performed by SPTPol.}
  \vspace{0.2cm}
  \begin{tabular}{|c | c | c | c|}
\hline
Surveys &$\alpha_{r=0}, N^{B}_{\ell}=0$& $\alpha_{r=0}$ & $\alpha_{r=0.12}$ \\ \hline \hline
SPTPol (31\%) &1.46& 1.27  & 1.08 \\ \hline
LSS-S2 +SPTPol (54\%) & 2.24 & 1.59 & 1.14  \\ \hline

@@

\begin{table}
\caption{$\alpha$: Improvements on $\sigma(r)$ due to delensing for S3 experiments. Here we define LSS-S3 as the combination of S2 LSS tracers (LSS-S2) and DESI. We also use a CMB S3 experiment for the lensing internal reconstruction. See \refsec{S3} for details.}
%The absolute error value of $\sigma(r)$ after delensing for $f_{sky} = $ is $\sigma(r) =$ 
\hspace{2.2cm}
\begin{tabular}{|c | c | c | c|}
\hline
Surveys & $\alpha_{r=0}, N^{B}_{\ell}=0$ & $\alpha_{r=0}$ & $\alpha_{r=0.12}$ \\ \hline \hline
DESI (10\%) &1.12 & 1.08  & 1.03 \\ \hline
LSS-S3 (43\%) &1.82& 1.42  & 1.12\\ \hline
CMB 3G (56\%) &2.38& 1.9  & 1.16 \\ \hline
LSS-S3+CMB (69\%)& 3.2 & 2.3& 1.21  \\ \hline
\end{tabular}
\label{tab:S3}
\end{table}
